---
title: "Tutorial: an interactive R shiny app visulaling K-means clustering of the Iris dataset"
author: "Zixuan Yu"
date: "2/9/2023"
output: md_document
---
#### Introduction 
The Iris dataset consists of 150 observations of iris flowers, each has four attributes: Sepal.Length, Sepal.Width, Petal.Length, and Petal.Width. There are three different species of iris in the dataset. The goal of this tutorial is to implement a Shiny R app that allows the user to cluster the samples based on the four attributes, and to visualize the results.  


#### Load the required libraries:  
```{r message=FALSE}
library(shiny)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(knitr)
```

#### preparing the data
load the data  
```{r}
data("iris")
```
Then we would extract the attributes to perform K-means clustering    
```{r}
df <- iris[,-5] # excluding the Species column
species <- iris$Species
```
Run PCA to reduce dimensionality     
```{r}
pca <- prcomp(df)
```

#### Performing K-means clustering
the custermized function for K-means clustering  
```{r}
kmeans_clustering <- function(df, K) {
        # Fit the k-means model
        clustering_fit <- kmeans(df, centers = K)
        
        # Cluster assignments
        cluster_assignments <- clustering_fit$cluster
        
        return(list(cluster_assignments, clustering_fit))
}
```


### the UI layour
```{r eval=FALSE}
# UI
ui <- fluidPage(
        titlePanel("K-Means Clustering of the Iris dataset"),
        
        fluidRow(
                #selecting number of clusters
                column(2,numericInput("K", "Number of Clusters", 
                                      value = 4, min = 1, max = 9,
                                      step = 1)),
                #clustering assignment displayed in plot
                column(9,plotOutput("plot"))
        ),
        fluidRow(column(3,tableOutput("table1")),
                 column(7,tableOutput("table2"))
        ),
        
        hr(),
        print("by Zixuan Yu   Feb 8th,2023") # add the footer
        
)
```




### The server components
We would use **K=3** to plot an sample scatter plot in this tutorial, but the final R Shiny App would allow user to change the value of K (from 1 to 9).   
  
  
#### the scatter plot  
```{r}
K = 3
cluster_results <- kmeans_clustering(pca$x[, 1:2], K)
# in the server function, you would need to wrap the right side with reactive ({ })

ggplot(data.frame(pca$x[, 1:2]), 
       aes(x = PC1, y = PC2, 
           color = species, # color by species
           shape = factor(cluster_results[[1]]))) +
                #shape representing cluster assignment
        geom_point(size = 3, show.legend=TRUE) + 
        scale_color_brewer(type = "qual", palette = "Set2") + 
        # change color palette
        scale_shape_manual("cluster",values = c(1:K))+
        ggtitle( paste("K-Means Clustering with K =", K))+
        theme_bw()
```
  
#### Table1 showing how many species in each cluster  
```{r}
 df_res <- iris # make a copy of the original data used 
df_res$cluster = cluster_results[[1]] # add cluster assignment to df_res
                df_res %>%
                        group_by(cluster,Species) %>% 
                        tally() %>%
        kable()
```

#### Table2 showing the summary statistics of each cluster  
```{r}
df_res %>% group_by(cluster) %>%
                        summarise(n = n(),
                                avg.Sepal.Length = mean(Sepal.Length),
                                avg.Sepal.Width = mean(Sepal.Width),
                                avg.Petal.Length = mean(Petal.Length),
                                avg.Petal.Width = mean(Petal.Width)) %>%
        kable()

```

#### Put the them together! The following chunk is an demonstration of the server function 
```{r eval=FALSE}
# Server
server <- function(input, output) {
        # Cluster the data
        cluster_results <- reactive({
                kmeans_clustering(pca$x[, 1:2], input$K)
        })
        
        # Plot the results
        output$plot <- renderPlot({
                # use PCA result to plot (for the purpose of dimensional reduction)
                ggplot(data.frame(pca$x[, 1:2]), aes(x = PC1, y = PC2, 
                                                     color = species, 
                                                        # color by species
                                                     shape = factor(cluster_results()[[1]]))) + 
                                                        # shape representing cluster assignment
                        geom_point(size = 3, show.legend=TRUE) + 
                        scale_color_brewer(type = "qual", palette = "Set2") + 
                                # change color palette
                        scale_shape_manual("cluster",values = c(1:input$K))+
                        ggtitle( paste("K-Means Clustering with K =", input$K))+
                        theme_bw()
        })
        
        df_res <- iris # make a copy of the original data used 
        
        # Table1 showing how many species in each cluster
        output$table1 <- renderTable({
                df_res$cluster = cluster_results()[[1]] # add cluster assignment to df_res
                df_res %>%
                        group_by(cluster,Species) %>% 
                        tally() # count how many Species in each cluster
        })
        
        # Table2 showing the summary statistics of each cluster
        output$table2 <- renderTable({
                df_res$cluster = cluster_results()[[1]] # add cluster assignment to df_res
                df_res %>% group_by(cluster) %>%
                        summarise(n = n(),
                                avg.Sepal.Length = mean(Sepal.Length),
                                avg.Sepal.Width = mean(Sepal.Width),
                                avg.Petal.Length = mean(Petal.Length),
                                avg.Petal.Width = mean(Petal.Width))

                })

}
```


#### Put the server and UI components together! 
```{r eval=FALSE}
shinyApp(ui, server)
```
